CF Registry Analysis - ROC

### 2025-11-19 Christine Chan/CF Registry/cf_registry_ROC_analysis.qmd
started piecewise regression for HbA1c project

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
library(childsds)
library(labelled)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/GitHub",
  "Windows" = "C:/Users/Tim/GitHub",
  "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

@@

    Vx809comb

@@

# Raw values for adults
roc_df$bmi_adult[roc_df$Age_YrEnd > 20] = NA

@@

@@

```{r}
# Plotting function
roc_plot <- function(roc_obj) {
  a <- ci.auc(roc_obj)
  a <- round(as.numeric(a[1:3]), 2)
  c <- coords(roc_obj, "best")
  plot_df <- coords(roc_obj)
  point_df <- plot_df[plot_df$threshold == as.numeric(c[1]), ]
  c <- round(as.numeric(c), 2)
  point_label <- paste0(
    "Threshold: ",
    c[1],
    "\n",
    "Specificity: ",
    c[2],
    "\n",
    "Sensitivity: ",
    c[3],
    "\n",
    "AUC: ",
    a[2],
    " (95%CI: ",
    a[1],
    ", ",
    a[3],
    ")"
  )
  p <- ggplot(plot_df, aes(x = 1 - specificity, y = sensitivity)) +
    geom_line() +
    geom_point(data = point_df) +
    geom_label_repel(
      data = point_df,
      aes(label = point_label),
      nudge_x = 0.25,
      segment.color = "grey50",
      size = 5
    ) +
    xlab("1-Specificity") +
    ylab("Sensitivity") +
    theme_bw(base_size = 20)
  return(p)
}
```

@@

## NGT vs. (AGT + CFRD) and (NGT + AGT) vs. CFRD

@@

1. NGT vs. (AGT + CFRD)

2. (NGT + AGT) vs. CFRD

```{r}
ngt_agt <- roc_df |>
  mutate(
    ogtt = factor(
      `OGTT 2-hour`,
      levels = c("<140", "[140,200)", ">=200"),
      labels = c("NGT", "AGT + CFRD", "AGT + CFRD")
    )
  )
agt_cfrd <- roc_df |>
  mutate(
    ogtt = factor(
      `OGTT 2-hour`,
      levels = c("<140", "[140,200)", ">=200"),
      labels = c("NGT + AGT", "NGT + AGT", "CFRD")
    )
  )
```

@@

#### NGT vs. (AGT + CFRD)

@@

#### (NGT + AGT) vs. CFRD

@@

##### NGT vs. (AGT + CFRD)

@@

##### (NGT + AGT) vs. CFRD

@@

@@

# FEV1 over time plot

```{r}
plot_df <- annual |>
  filter(
    !is.na(cfrd_status_annual),
    Age_YrEnd >= 10,
    !cfrd_status_annual %in% c(4, 5)
  ) |>
  group_by(eDWID) |>
  mutate(
    Diagnosis = max(cfrd_status_annual),
    `Years to CFRD or Last Visit` = case_when(
      Diagnosis == 3 ~
        Age_YrEnd -
        first(Age_YrEnd[cfrd_status_annual == 3]),
      .default = Age_YrEnd - max(Age_YrEnd, na.rm = T)
    )
  )
plot_df$Diagnosis <- factor(
  plot_df$Diagnosis,
  levels = 1:3,
  labels = c("NGT", "AGT", "CFRD")
)
ggplot(
  plot_df,
  aes(
    y = A_FEV1_pct_predicted,
    x = `Years to CFRD or Last Visit`,
    color = Diagnosis,
    group = Diagnosis
  )
) +
  xlim(c(-10, 10)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_smooth(formula = y ~ s(x, bs = "cr")) +
  theme_bw()
```

>>>>

```{r setup}
#| include: false
library(tidyverse)
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
library(childsds)
library(labelled)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/GitHub",
  "Windows" = "C:/Users/Tim/GitHub",
  "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

@@

    twohour_bloodglucose

@@

@@

# Raw values for adults, percentiles for peds
roc_df$bmi_adult[roc_df$Age_YrEnd <= 20] = NA
roc_df$bmi_perc = roc_df$bmi_perc
roc_df$bmi_perc[roc_df$Age_YrEnd > 20] = NA

@@

```{r}
# Plotting function
roc_plot <- function(roc_obj) {
  a <- ci.auc(roc_obj)
  a <- round(as.numeric(a[1:3]), 2)
  c <- coords(roc_obj, "best")
  mod_df <- coords(roc_obj)
  mod_df$one_m_spec = 1 - mod_df$specificity
  point_df <- mod_df[mod_df$threshold == as.numeric(c[1]), ]
  c <- round(as.numeric(c), 2)
  point_label <- paste0(
    "Threshold: ",
    c[1],
    "\n",
    "Specificity: ",
    c[2],
    "\n",
    "Sensitivity: ",
    c[3],
    "\n",
    "AUC: ",
    a[2],
    " (95%CI: ",
    a[1],
    ", ",
    a[3],
    ")"
  )
  p <- ggplot(mod_df, aes(x = one_m_spec, y = sensitivity)) +
    geom_line() +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    geom_point(data = point_df) +
    geom_label_repel(
      data = point_df,
      aes(label = point_label),
      nudge_x = 0.5,
      nudge_y = -0.5,
      segment.color = "grey50",
      size = 5
    ) +
    xlab("1-Specificity") +
    ylab("Sensitivity") +
    theme_bw(base_size = 20)
  return(p)
}
```

@@

## NGT vs. AGT and AGT vs. CFRD

@@

1. NGT vs. AGT

2. AGT vs. CFRD

```{r}
ngt_agt <- roc_df |> filter(`OGTT 2-hour` != ">=200")
ngt_agt$ogtt <- droplevels(ngt_agt$`OGTT 2-hour`)
agt_cfrd <- roc_df |> filter(`OGTT 2-hour` != "<140")
agt_cfrd$ogtt <- droplevels(agt_cfrd$`OGTT 2-hour`)
```

@@

#### NGT vs. AGT

@@

#### AGT vs. CFRD

@@

##### NGT vs. AGT

@@

##### AGT vs. CFRD

@@

##### $\geq$ 200
Bootstrap replicates found more than one "best" threshold, so the non-bootstrapped threshold is reported first, and a random bootstrapped "best" interval second.

@@

# Diagnosis comparisons for HbA1c registry paper

```{r}
#| warning: false
#| message: false
library(Hmisc)
library(lmerTest)
# Import data and codebook
data_dict <- readxl::read_excel("./Data_Raw/Copy of Codes for CFFPR_2018.xlsx")
data_dict$var <- paste(data_dict$field_name, data_dict$code_value, sep = "_")
# Remove those with T1D or T2D
annual <- annual %>% arrange(eDWID, ReviewYear)
annual <- annual[!annual$cfrd_status_annual %in% c(4, 5), ]
annual$cfrd_status_annual <- factor(
  annual$cfrd_status_annual,
  labels = c("CFRD-", "CFRD-", "CFRD+")
# Get mutations
demogs <- read.csv("./Data_Raw/CFF19_DemogCFDiag_Del1.csv")
  "Native Hawaiian or Other Pacific ",
demogs$Race <- apply(demogs, 1, function(r) {
  if (length(w) > 1) {
    return("More than one race")
    return(races[w])
demogs$Race <- factor(
  demogs$Race,
    "Native Hawaiian or Other Pacific ",
demogs$Hispanicrace <- factor(
  demogs$Hispanicrace,
demogs <- demogs %>% select(eDWID, Gender, MutClass, Race, Hispanicrace)
annual <- left_join(annual, demogs, by = join_by(eDWID))
# Variable formatting
# If A_supplement_feeding is 0, then A_feeding2 is also 0 because they are not
# on any supplemental feeding
# Include all kinds of tube feeds
annual$A_feeding2[annual$A_supplement_feeding == 0] <- 0
annual$A_feeding3[annual$A_supplement_feeding == 0] <- 0
annual$A_feeding4[annual$A_supplement_feeding == 0] <- 0
tube_fields <- c("A_feeding2", "A_feeding3", "A_feeding4")
annual$tube_feeding <-
  apply(annual[, tube_fields], 1, function(r) {
    if (all(is.na(r))) {
      return(NA)
      return(max(as.numeric(r), na.rm = T))
# Liver disease
liver_fields <- c("A_hepatobiliary1_3", paste0("A_hepatobiliary2_", 1:4))
annual$liver_disease <-
  apply(annual[, liver_fields], 1, function(r) {
    if (all(is.na(r))) {
      return(NA)
      return(max(as.numeric(r), na.rm = T))
# Get BMI categories
annual$bmi <- annual$A_weight / ((annual$A_height / 100)^2)
annual$bmi <- coalesce(annual$A_bmivalue, annual$bmi)
annual$bmi_perc_cdc <- sds(
  annual$bmi,
  age = annual$Age_YrEnd,
  sex = annual$Gender,
  ref = cdc.ref,
  type = "perc"
annual$bmi_perc_who <- sds(
  annual$bmi,
  age = annual$Age_YrEnd,
  sex = annual$Gender,
  ref = who.ref,
  type = "perc"
annual$bmi_perc <- coalesce(annual$bmi_perc_cdc, annual$bmi_perc_who)
annual$bmi_cat <- apply(annual, 1, function(r) {
annual$bmi_cat <- relevel(annual$bmi_cat, ref = "Target/Goal")
# Raw values for adults, percentiles for peds
annual$bmi_adult = annual$bmi
annual$bmi_adult[annual$Age_YrEnd <= 20] = NA
annual$bmi_perc = annual$bmi_perc
annual$bmi_perc[annual$Age_YrEnd > 20] = NA
var_label(annual$bmi) = "BMI (all participants)"
var_label(annual$bmi_perc) = "BMI percentile (age <= 20)"
var_label(annual$bmi_adult) = "BMI value (age > 20)"
var_label(annual$bmi_cat) = "BMI category"
# Format columns
annual$MutClass <- factor(
  annual$MutClass,
  levels = c("1-3", "4-5", "Oth", "N/A"),
  labels = c("1-3", "4-5", "Oth", "Unknown")
annual$NumPulmExacerbation <- cut(
  annual$NumPulmExacerbation,
  labels = c("0", "1", "2", ">=3"),
  breaks = c(0, 1, 2, 3, Inf),
  right = F
# Age groups 10 - 18 inclusive, 19 - 30 inclusive, >= 31
annual$age_group <- cut(
  annual$Age_YrEnd,
  breaks = c(-Inf, 10, 19, 31, Inf),
  right = F
# Bugs
bugs <- c(
  "A_pseudomonasaeruginosa",
  "A_fungalyeast1",
  "A_pulmonarycomplications1",
  "A_mycocultureresults3",
  "A_cultureresults3"
# Yes/No variables
yn_vars <- c(
  "A_corticosteroids1",
  "A_feeding2",
  "A_Vx770",
  "A_VX809comb",
  bugs,
  "tube_feeding",
  "liver_disease"
annual[, yn_vars] <- lapply(
  annual[, yn_vars],
  factor,
annual$A_IsOnEnzymes <- factor(
  annual$A_IsOnEnzymes,
  levels = c(1, 0),
  labels = c(" Yes", " No")
# FEV1 to categorical
annual$ppFEV1_cat <- cut(
  annual$A_FEV1_pct_predicted,
  breaks = c(0, 40, 80, 100, Inf),
  right = F
annual$ppFEV1_cat <- factor(
  annual$ppFEV1_cat,
  levels = c("[100,Inf)", "[80,100)", "[40,80)", "[0,40)")
# Males to reference group
annual$Gender <- factor(annual$Gender, levels = c("M", "F"))
# Hispanic no as reference
annual$Hispanicrace <- factor(annual$Hispanicrace, levels = c("No", "Yes"))
# Get first year of CFRD
first_cfrd <- annual %>%
  filter(cfrd_status_annual == "CFRD+") %>%
  group_by(eDWID) %>%
  summarise(first_cfrd = min(ReviewYear))
annual <- left_join(annual, first_cfrd, by = join_by(eDWID))
# Year prior to first CFRD in registry - assume negative for CFRD. Assume all
# years following are CFRD+
annual$CFRD <- NA
annual$CFRD[annual$ReviewYear < annual$first_cfrd] <- "CFRD-"
annual$CFRD[annual$ReviewYear >= annual$first_cfrd] <- "CFRD+"
annual$CFRD[is.na(annual$first_cfrd)] <- "CFRD-"
annual$CFRD <- factor(annual$CFRD, levels = c("CFRD-", "CFRD+"))
# Add formatted columns for survival analysis
annual$tstop <- ymd(paste0(annual$ReviewYear, "-12-31"))
annual$tstart <- ymd(paste0(annual$ReviewYear, "-01-01"))
annual <- annual %>%
  group_by(eDWID) %>%
  mutate(entry_date = tstart[1]) %>%
  ungroup()
# Remove those under 10
annual$age_group <- relevel(annual$age_group, ref = "[10,19)")
annual <- annual %>% filter(Age_YrEnd != 0, Age_YrEnd >= 10)
annual$age_group <- droplevels(annual$age_group)
# List of covariates
vars <- c(
  "age_group",
  "Gender",
  "bmi_perc",
  "bmi_cat",
  "MutClass",
  "Race",
  "Hispanicrace",
  "A_IsOnEnzymes",
  "ppFEV1_cat",
  "tube_feeding",
  "NumPulmExacerbation",
  "A_pseudomonasaeruginosa",
  "A_fungalyeast1",
  "A_mycocultureresults3",
  "A_corticosteroids1",
  "cfrd_status_annual",
  "A_Vx770",
  "A_VX809comb",
  "liver_disease"
# Nice labels for plots and tables
labels <- list(
  cfrd_status_annual = "CFRD Status",
  MutClass = "Mutation class",
  Race = "Race",
  Hispanicrace = "Hispanic",
  Gender = "Sex",
  A_IsOnEnzymes = "Pancreatic enzyme use",
  ppFEV1_cat = "FEV1 percent predicted category",
  A_FEV1_pct_predicted = "FEV1 percent predicted",
  A_FVC_pct_predicted = "FVC percent predicted",
  tube_feeding = "Enteric feedings",
  liver_disease = "Liver disease",
  A_corticosteroids1 = "Use of corticosteroids",
  NumPulmExacerbation = "Admissions for PEx",
  age_group = "Age group",
  height_perc = "Height percentile",
  A_Vx770 = "Use of CFTR modulator Vx770",
  A_VX809comb = "Use of CFTR modulator Vx809 combination",
  A_pseudomonasaeruginosa = "Pseudomonas aeruginosa",
  A_fungalyeast1 = "Aspergillus (any species)",
  A_mycocultureresults3 = "Microorganisms in myco culture",
  A_cultureresults3 = "Microorganisms in culture",
  eDWID = "ID",
  tstart = "Interval start",
  tstop = "Interval stop",
  Age_YrEnd = "Age"
label(annual[, which(colnames(annual) %in% names(labels))]) <-
  labels[colnames(annual)[which(colnames(annual) %in% names(labels))]]
# Per reviewer, out of X individuals with a diagnosis of CFRD, X number had an
# OGTT (and X had an A1c, but no OGTT) that year or the year prior.
# This is super slow but idc idc
cfrd_only <- apply(first_cfrd, 1, function(r) {
  id <- as.numeric(r["eDWID"])
  year <- as.numeric(r["first_cfrd"])
  d <- annual[annual$eDWID == id & annual$ReviewYear %in% c(year, year - 1), ]
  a1c <- as.numeric(last(d["A_hgba1c"], na_rm = TRUE))
  ogtt0hr <- as.numeric(last(d["A_ogttfast_bloodglucose"], na_rm = TRUE))
  ogtt2hr <- as.numeric(last(d["A_twohour_bloodglucose"], na_rm = TRUE))
  return(c(id, year, a1c, ogtt0hr, ogtt2hr))
cfrd_only <- data.frame(t(cfrd_only))
colnames(cfrd_only) <- c("id", "year", "a1c", "ogtt0hr", "ogtt2hr")
cfrd_only <- left_join(
  cfrd_only,
  annual %>% mutate(eDWID = as.numeric(eDWID)),
  by = join_by(id == eDWID, year == ReviewYear)
cfrd_only$a1c_ogtt <- apply(cfrd_only, 1, function(r) {
  not_missing <- paste(
    which(!is.na(r[c("a1c", "ogtt0hr", "ogtt2hr")])),
    collapse = "_"
  return(not_missing)
cfrd_only$a1c_ogtt <- factor(
  cfrd_only$a1c_ogtt,
  levels = c("", "1", "1_2", "1_2_3", "1_3", "2", "2_3", "3"),
    "Neither",
    "HbA1c Only",
    "OGTT",
    "OGTT",
    "OGTT",
    "OGTT",
    "OGTT",
    "OGTT"
cfrd_only %>%
    a1c_ogtt,
    A_weight,
    A_height,
    bmi_adult,
    bmi_cat,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3
  ) %>%
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  add_p(test = list(all_continuous() ~ "oneway.test")) %>%
  modify_caption("**All Groups**")
cfrd_only %>%
  filter(a1c_ogtt %in% c("OGTT", "HbA1c Only")) %>%
  mutate(a1c_ogtt = droplevels(a1c_ogtt)) %>%
    a1c_ogtt,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3
  ) %>%
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  add_p(test = list(all_continuous() ~ "oneway.test")) %>%
  modify_caption("**OGTT vs. HbA1c Only**")

## FEV1 trends by diagnosis method

### LOESS-smoothed plot

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
mod_df = cfrd_only %>%
  select(id, a1c_ogtt) %>%
  rename(eDWID = id)
mod_df = left_join(
  annual %>% mutate(eDWID = as.numeric(eDWID)),
  mod_df,
  by = join_by(eDWID)
)
mod_df = mod_df %>%
  group_by(eDWID) %>%
  mutate(years_to_cfrd = ReviewYear - first_cfrd)
var_label(mod_df$years_to_cfrd) = "Years to CFRD"
ggplot(
  mod_df %>% filter(!is.na(a1c_ogtt)),
  aes(
    y = A_FEV1_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlim(c(-10, 10)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_smooth(formula = y ~ s(x, bs = "cr")) +
  theme_bw() +
  theme(legend.title = element_blank())
```

### Piecewise linear model results

```{r}
# Fit model
mod_df$x = mod_df$years_to_cfrd * (mod_df$years_to_cfrd >= 0)
label(mod_df$x) = "Change in Slope Post-CFRD"
label(mod_df$years_to_cfrd) = "Slope Pre-CFRD"
label(mod_df$a1c_ogtt) = "Diagnosis Method"
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    x * a1c_ogtt +
    (1 | eDWID),
  data = mod_df
)
# Population-averaged plot

# Model table
m |> tbl_regression()
```
